---
title: "COVID 19 Analysis"
output: html_document
date: "2025-05-13"
---

### 0) Install Necessary Packages

```{r setup}
# use install.packages("<insert pakage name>") command in console if any of the following packages are not already installed
library(tidyverse)
library(scales)
library(car)
```

### 1) Read in CSV Data from GitHub

For this analysis, I will download global data for cases, deaths, and recoveries, as well as US data for cases and deaths.

```{r read_csvs, message=FALSE}
# get CSVs from GitHub
url_start <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv", "time_series_covid19_recovered_global.csv", "time_series_covid19_confirmed_US.csv", "time_series_covid19_deaths_US.csv")
urls <- str_c(url_start, file_names)

# read data from CSVs
global_cases_unmod <- read_csv(urls[1])
global_deaths_unmod <- read_csv(urls[2])
global_recoveries_unmod <- read_csv(urls[3])
us_cases_unmod <- read_csv(urls[4])
us_deaths_unmod <- read_csv(urls[5])
```

### 2) Clean Up Global Data

Part 1: Reorganize data with one date per row, remove latitude and longitude (unneccessary for my analysis), and change new "Date" variable to date type

```{r clean_global_data}
# create new datasets with modified data so originals are preserved
global_cases <- global_cases_unmod %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long),
               names_to = "Date",
               values_to = "Cases") %>%
  select(-c(Lat, Long)) %>%
  mutate(Date = mdy(Date))

global_deaths <- global_deaths_unmod %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long),
               names_to = "Date",
               values_to = "Deaths") %>%
  select(-c(Lat, Long)) %>%
  mutate(Date = mdy(Date))

global_recoveries <- global_recoveries_unmod %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long),
               names_to = "Date",
               values_to = "Recoveries") %>%
  select(-c(Lat, Long)) %>%
  mutate(Date = mdy(Date))
```

Part 2: Transform global data by combining cases, deaths, and recoveries

```{r combine_global}
global <- global_cases %>%
  full_join(global_deaths) %>%
  full_join(global_recoveries) %>%
  rename(Province_State = 'Province/State',
         Country_Region = 'Country/Region')

# show summary
summary(global)
```

### 3) Clean Up US Data

Part 1: Reorganize data with one date per row, remove latitude and longitude (unnecessary for my analysis), and change new "Date" variable to date type


```{r clean_us_data}
# create new datasets with modified data so originals are preserved
us_cases <- us_cases_unmod %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "Date",
               values_to = "Cases") %>%
  select(Admin2:Cases) %>%
  select(-c(Lat, Long_)) %>%
  mutate(Date = mdy(Date))

# there is an extra "Population" variable in the deaths dataset
us_deaths <- us_deaths_unmod %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "Date",
               values_to = "Deaths") %>%
  select(Admin2:Deaths) %>%
  select(-c(Lat, Long_)) %>%
  mutate(Date = mdy(Date))
```

Part 2: Transform US data by combining cases and deaths

```{r combine_us}
us <- us_cases %>%
  full_join(us_deaths) %>%
  # rename "Admin2" variable to "County" for easier identification
  rename(County = Admin2) %>%
  # reorder "Deaths" and "Population" variables so that the former comes right after the "Cases" variable
  relocate(Deaths, .before = Population)

#show summary
summary(us)

# at time of writing, the following commands show the same three dates with negative numbers of cases and deaths:
us %>% filter(Cases < 0)
us %>% filter(Deaths < 0)

# modify dataset to remove these possible errors
us <- us %>%
  filter(Cases >= 0)
```

### 4) Modify Global Data for Easier Comparison with US Data

Part 1: Create new "Combined_Key" variable out of "Province_State" and "Country_Region" variables

```{r combine_location_variables}
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
```

Part 2: Add population data from another dataset

```{r add_global_population}
# import population information from another csv dataset
global_pop <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv", show_col_types = FALSE) %>%
  select(Province_State, Country_Region, Population)

# combine with "global" dataset
global_w_pop <- global %>%
  left_join(global_pop, by = c("Province_State", "Country_Region")) %>%
  select(Province_State, Country_Region, Combined_Key, Date, Cases, Deaths, Recoveries, Population)

# show summary
summary(global_w_pop)
```

### 5) How is global COVID 19 data predicted by case rate and recovery rate?

In the lecture, we walked through creating a model that predicted death rate based on case rate. We saw some correlation, but could tell other factors affected the death rate. For global data, we have additional data regarding case recoveries. I will implement a model to see if recovery rates alongside case rates affect the prediction of death rates per country.

Part 1: Create new dataset with total cases, deaths, and recoveries per country, along with associated rates based on population

```{r create_model_dataset}
# transform data to sum Cases/Deaths/Recoveries/Pop for all Province_State's within each Country_Region for each Date
global_mod_data <- global_w_pop %>%
  group_by(Country_Region, Date) %>%
  summarize(Cases = sum(Cases),
            Deaths = sum(Deaths),
            Recoveries = sum(Recoveries),
            Population = sum(Population)) %>%
  # transform data again by finding max Cases/Deaths/Recoveries/Pop for each country out of all days
  group_by(Country_Region) %>%
  summarize(Cases = max(Cases),
            Deaths = max(Deaths),
            Recoveries = max(Recoveries),
            Population = max(Population),
            Case_Rate = Cases / Population,
            Death_Rate = Deaths / Population,
            Recovery_Rate = Recoveries / Population) %>%
  # remove rows with 0 Pop (which will affect rates) or 0 Cases (which should equate to no Deaths)
  filter(Population > 0, Cases > 0) %>%
  ungroup()
```

Part 2a: To test our model compared the the one using US data from the lecture, first create and visualize a model that looks at death rates as a function of just case rates

```{r case_model}
global_mod <- lm(Death_Rate ~ Case_Rate, data = global_mod_data)
# show summary of model
summary(global_mod)

# add predictions variable to dataset, based on model
global_mod_data$Predictions = predict(global_mod)
```

Part 2b: We'll visualize it two ways, first with a plot similar to the one seen in lecture

```{r vis_mod_1}
# first plot
global_mod_data %>% ggplot() +
  geom_point(aes(x = Case_Rate, y = Death_Rate),
             color = "blue") +
  geom_point(aes(x = Case_Rate, y = Predictions),
             color = "red") +
  labs(title = "Actual vs. Predicted Death Rate", x = "Case Rate", y = "Death Rate") +
  scale_x_continuous(labels = label_percent(accuracy = 0.1)) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  theme_minimal()
```

Part 2c: The second plot evaluates a perfect prediction of the model against the real data

```{r vis_mod_2}
ggplot(global_mod_data, aes(x = Death_Rate, y = Predictions)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Actual vs. Predicted Death Rate", x = "Actual Death Rate", y = "Predicted Death Rate") +
  scale_x_continuous(labels = label_percent(accuracy = 0.1)) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  theme_minimal()
```

Part 3a: Now introduce recovery rates as an additional predictive factor in the model

```{r case_and_rec_model}
global_mod <- lm(Death_Rate ~ Case_Rate + Recovery_Rate, data = global_mod_data)
# show summary of model
summary(global_mod)

# add predictions variable to dataset, based on model
global_mod_data$Predictions = predict(global_mod)
```

Part 3b: Check for multicollinearity between case rate and recovery rate

```{r multicoll_check}
vif(global_mod)
```

At the time of this writing, the VIF Factors for both "Case_Rate" and "Recovery_Rate" are under 1.4. Multicollinearity is a topic I am not too familiar with, but based on these sources the VIF Factor for this model does not appear to be dangerously large:
https://pmc.ncbi.nlm.nih.gov/articles/PMC6900425/;
https://www.geeksforgeeks.org/multicollinearity-in-data/;
https://online.stat.psu.edu/stat462/node/180/

Part 3c: First plot with new model

```{r vis_mod_3}
# first plot
global_mod_data %>% ggplot() +
  geom_point(aes(x = Case_Rate, y = Death_Rate),
             color = "darkgreen") +
  geom_point(aes(x = Case_Rate, y = Predictions),
             color = "darkorange") +
  labs(title = "Actual vs. Predicted Death Rate", x = "Case Rate", y = "Death Rate") +
  scale_x_continuous(labels = label_percent(accuracy = 0.1)) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  theme_minimal()
```

Part 3d: Second plot with new model

```{r vis_mod_4}
ggplot(global_mod_data, aes(x = Death_Rate, y = Predictions)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, color = "darkorange") +
  labs(title = "Actual vs. Predicted Death Rate", x = "Actual Death Rate", y = "Predicted Death Rate") +
  scale_x_continuous(labels = label_percent(accuracy = 0.1)) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  theme_minimal()
```

Analysis: The first graph for the new dataset does appear to show a closer correlation with at between predictions and actuals for at least some data points, although both graphs make it clear that additional factors affect death rates. I might have some bias in guessing that countries with higher recovery rates will have lower death rates, but that might not necessarily be the case. As shown below, the country of Montenegro has the third highest recovery rate but also the sixth highest death rate.

```{r max_recovery_and_death_rates}
global_mod_data %>% slice_max(Recovery_Rate, n = 10) %>%
  select(Country_Region, Recovery_Rate, Death_Rate, everything())

global_mod_data %>% slice_max(Death_Rate, n = 10) %>%
  select(Country_Region, Recovery_Rate, Death_Rate, everything())
```

### 6) US States with Highest and Lowest Death Rates

For the five US states with the highest overall death rates, I want to find out how the death rate for each of them changes over time. I will then compare that with a similar graph for the five US states with the lowest overall death rates.

Part 1: Calculate death rate over time, grouping by state

```{r rates_over_time}
us_time_data <- us %>%
  group_by(Province_State, Country_Region, Date) %>%
  summarize(Cases = sum(Cases),
            Deaths = sum(Deaths),
            Population = sum(Population),
            Death_Rate = Deaths / Population) %>%
  filter(Population > 0) %>%
  ungroup()
```

Part 2: Calculate overall deaths rates per state

```{r overall_rates}
us_state_data <- us_time_data %>%
  group_by(Province_State) %>%
  summarize(Cases = max(Cases),
            Deaths = max(Deaths),
            Population = max(Population),
            Death_Rate = Deaths / Population) %>%
  ungroup()
```

Part 3: Identify the five states with the highest overall death rates and the five states with the lowest overall death rates

```{r identify_high_and_low}
# five states with highest overall death rates
us_state_data %>% slice_max(Death_Rate, n = 5)

# five states with lowest overall death rates
us_state_data %>% slice_min(Death_Rate, n = 5)
```

At the time of this writing, the five states with the highest overall death rates are (in order from greatest to least):
1. Arizona
2. Oklahoma
3. Mississippi
4. West Virginia
5. New Mexico

Meanwhile, the five states with the lowest overall death rates at the time of this writing are (in order from least to greatest):
1. American Samoa
2. Northern Mariana Islands
3. Virgin Islands
4. Hawaii
5. Vermont

Part 4a: Visualize change in death rate for five states with the overall highest death rates over time

```{r vis_highest}
# get list of just five states with highest overall death rate
max_states <- us_state_data %>%
  slice_max(Death_Rate, n = 5)

#visualize death rate change over time
us_time_data %>%
  filter(Province_State %in% max_states$Province_State) %>%
  ggplot(aes(x = Date, y = Death_Rate, color = Province_State)) +
  geom_line(linewidth = 0.5) +
  labs(title = "Death Rate Over Time for Five States with Highest Overall Death Rates", x = "Date", y = "Death Rate", color = "State") +
  theme_minimal() +
  scale_y_continuous(labels = label_percent(accuracy = 0.1))
```

Part 4b: Visualize change in death rate for five states with the overall lowest death rates over time

```{r vis_lowest}
# get list of just five states with lowest overall death rate
min_states <- us_state_data %>%
  slice_min(Death_Rate, n = 5)

#visualize death rate change over time
us_time_data %>%
  filter(Province_State %in% min_states$Province_State) %>%
  ggplot(aes(x = Date, y = Death_Rate, color = Province_State)) +
  geom_line(linewidth = 0.5) +
  labs(title = "Death Rate Over Time for Five States with Lowest Overall Death Rates", x = "Date", y = "Death Rate", color = "State") +
  theme_minimal() +
  scale_y_continuous(labels = label_percent(accuracy = 0.1))
```

Analysis: I was curious to see if the states with the highest overall death rates at any point had death rates lower than those of the states with the lowest overall death rates. Based on the data I can see at the time of this writing, that doesn't appear to be the case. Something I didn't consider when I conceived of this analysis was that many of the lowest states (or territories) would be islands, which could have had a significant impact on how and when COVID spread in those locations. My preconceived bias was that certain states may have performed better or worse in terms of death rates depending on whether or not those states took proactive steps to contain the spread of the illness and/or distribute vaccines.

### 7) Conclusion

In this report, I attempted to analyze (1) how a combination of case rate and recovery rate could predict death rate using the global dataset, and (2) what were the US states/territories with the highest and lowest death rates.

My analysis of the former identified a low multicollinearity between case rate and recovery rate, and a closer prediction of death rate based on both factors compared to that based on just case rate. However, there are still clearly additional factors that played an important role in death rates across the globe. I have a guess that countries with higher recovery rates may have lower death rates, but it is important to not treat that as an assumption that might unduly affect how I display my work.

My analysis of the second question revealed that islands tended to be among the lowest death rate states/territories in the US. Originally I had thought the outcome might be more determined by stay-at-home and vaccination policies, and it was important for me to remain mindful of that bias while designing my functions and visualizations.